Depolarization of interacting NV$^-$ centers at low magnetic fields

### 2022-07-12 Mes papiers/Low field CR/V3/SI/Low_field_CR_SI.tex
la tete de tes morts

\section{NV$^-$ ground state Hamiltonian under low magnetic field}
\label{sec Hamiltonian}
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_ESR}
\caption{ODMR measurement (a) under zero magnetic field, (b) under a magnetic field of $~ 50 G$ when zooming on a single NV class}
\label{ESR_single_spin}
\end{figure}
Under zero external magnetic field, there are three possible processes that can cause a splitting of the $\{\ket{+1},\ket{-1}\}$ states : local electric fields, crystal strain and local magnetic fields. Of these three causes, only the electric field can explain the shape of the ODMR line that is observed under zero external magnetic field (Fig. \ref{ESR_single_spin}a) \cite{mittiga2018imaging}.

@@

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/map_etats_propres}
\caption{Numerical simulations showing the closeness of the Hamiltonian second excited state $\ket{e}$ with the states $\ket{+1}$ and $\ket{+}$, as a function of the magnetic field amplitude and angle $\theta$ with respect to the NV axis. A value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen}
\label{map etats propres}
\end{figure}

In our model for  dipole-dipole coupling under small magnetic fields, we will therefore neglect the contribution of the strain, local magnetic fields and longitudinal electric fields.  We also do not take into account the hyper-fine structure of the NV center because of the large inhomogeneous broadening of the transitions in all our samples (Fig. \ref{ESR_single_spin}b).
We will then consider the following spin Hamiltonian for the NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_s/h=D S_z^2 + \gamma_e \bm{B}_{\rm ext} \cdot \bm{S}+ d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
Where $D=2.87\ \rm GHz$ is the zero field splitting and $\gamma_e=2.8\ \rm MHz/G$ the gyromagnetic ratio of the electron.

@@

\section{Samples}
%Note pour moi : j'ai inversé adamas 1 et 2 parraport aux noms des dossiers dans 20220324
Here are the various samples used in this study :
\begin{itemize}
\item \textbf{HPHT-150-1} : A 150 $\mu$m HPHT 1b diamond irradiated and annealed to reach a concentration [NV$^-$] $\approx 3\ \rm ppm$. This sample was bought from Adamas Nanotechnology (MDNV150umHi). It is used in Fig. 2 (b), 2(d) and 3 of the main text, as well as Fig. \ref{ESR_single_spin}, \ref{T1_fits} and \ref{alphas} of the SI.
\item \textbf{HPHT-150-2} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. 4 of the main text and Fig. \ref{largeur_fluct} (c) of the SI.
\item \textbf{HPHT-150-3} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. \ref{Pola} of the SI.
\item \textbf{HPHT-15-1}  : A 15 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV15umHi). This sample is used in Fig. 5 of the main text.
\item \textbf{HPHT-15-2}  : Another sample from the same batch as HPHT-15-1. This sample is used in Fig. \ref{121 VS 22 fig} and \ref{Various ODMR} of the SI.
\item \textbf{HPHT-1-1}  : A 1 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV1umHi). This sample is used to perform a sensitivity measurement similar to the one done in Fig. 5 of the main text.
\item \textbf{CVD-PPM} : A CVD bulk diamond described in \citep{TALLAIRE2020421}. It has been irradiated and annealed and contains [NV$^-$] $\approx 4\ \rm ppm$. This sample is used in Fig. \ref{Alignment} of the SI.
\item \textbf{CVD-PPB} : Another CVD bulk diamond. This one has not been irradiated or annealed and therefore contains [NV$^-$] $\approx 50\ \rm ppb$. This sample is used in Fig. 2(a) of the main text.

@@

\section{Experimental Setup}
\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/shema_exp}
\caption{Experimental Setup}
\label{setup}
\end{figure}
Fig. \ref{setup} shows the general purpose experimental setup used for all the experiments presented in this article.

@@

We did not use magnetic shielding to protect from the earth magnetic field since the splitting due to  $B_{\rm earth}$ is lower than the splitting due to the local electric field for the samples used here : $ \gamma_e B_{\rm earth} \approx 1.5\ \rm MHz < d_\perp E_\perp \approx 4\ \rm MHz$. 
When possible, the electromagnet was roughly aligned along the earth magnetic field to better compensate it.
\section{Experimental details}
\subsection{$T_1$ fitting Protocol}
\begin{figure}
\includegraphics[width=0.8\textwidth]{Figures_SI/Fig_T1_combined}
\caption{$T_1$ measurement with purely exponential and purely stretched exponential fits (a) in zero magnetic field (b) in non-zero magnetic field}
\label{T1_fits}
\end{figure}
\begin{figure}[h]
\includegraphics[width=0.45\textwidth]{Figures_SI/fig_alphas}
\caption{Best stretch factor $\beta$ for a $T_1$ fit of the form $f(\tau)=A \exp(-(\frac{\tau}{T_1})^\beta)$ as a function of an arbitrarely oriented magnetic field amplitude}
\label{alphas}
\end{figure}

@@

Fig. \ref{T1_fits} shows the result of a lifetime measurement, measured with the subtraction protocol described in the main text, for zero and non-zero magnetic field. The result is fitted with exponential and stretched-exponential (with a stretch factor $\beta=0.5$) functions. We can see that for the shorter lifetime (zero field), the measurement follows more closely the stretched-exponential profile, while the longer lifetime is closer to the exponential one (except at very short times).

@@

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/overlap}
\caption{Illustration of the spectral overlap for two gaussian spectra}
\label{overlap}
\end{figure}

@@

As illustrated in Fig. \ref{overlap}, we define the spectral overlap $S(\Delta \nu)$ between two spins of spectral response $S_1(\nu)$ and $S_2(\nu)$, centered respectively on the frequencies $\nu_1$ and $\nu_2$ where $\Delta \nu = \nu_2-\nu_1$ as :
\begin{equation}
S(\Delta \nu)=\int S_1(\nu, \nu_1)S_2(\nu, \nu_2) d\nu
\end{equation}
In order to approximate the spectral overlap in our experiment, we will consider the analytical solution in the Gaussian and Lorentzian case :

\begin{itemize}
\item For two gaussians of standard deviation $\sigma$, the spectral overlap as a function of the detuning $\Delta \nu = \nu_1-\nu_2$ is itself a gaussian of standard deviation $\sigma'=\sqrt{2} \sigma$. :
\begin{align*}
S(\Delta \nu)&\propto \int \exp(-\frac{(\nu-\nu_1)^2}{2\sigma^2})\exp(-\frac{(\nu-\nu_2)^2}{2\sigma^2}) d\nu \\
&\propto\exp(-\frac{(\Delta \nu)^2}{4\sigma^2})
\end{align*}

\item For two Lorentzian profile with width $\sigma$, the overlap function is itself a Lorentizan with width $\sigma'=2\sigma$ :
\begin{align*}
S(\Delta \nu)&\propto \int \frac{1}{1+ \frac{(\nu-\nu_1)^2}{\sigma^2}}\cdot \frac{1}{1+ \frac{(\nu-\nu_2)^2}{\sigma^2}} d\nu \\
&\propto\frac{1}{1+ \frac{(\Delta \nu)^2}{4\sigma^2}}
\end{align*}
\end{itemize}

@@

In particular, we did not observe the apparition of an anti-line inside the main dip, unlike what was observed in the two previously cited work. We expect that the main reason behind the different behaviors is that we seem to observe far greater spin depolarization in zero-field, which we attribute to dipole-dipole coupling. This effect could hide smaller effects such as the one related to the laser polarization.
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_Pola}
\caption{Effect of the polarization of the incident laser. (a) Photoluminescence of sample HPHT-150-3 as a function of arbitrarely oriented magnetic field amplitude for various polarization angle. (b) Demodulated PL in the same conditions.}
\label{Pola}
\end{figure}
\subsection{Alignment of the magnetic field along [100] axis}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{Figures_SI/alignement}
\caption{PL of sample CVD-PPM as a function of the external magnetic field amplitude for different field misalignment with the [100] axis.}
\label{Alignment}
\end{figure}
We claim that the decrease in PL at low magnetic field in Fig. 3 (c)(ii) of the main text is due to the local electric field and the double flip processes. One other possibility would be a misalignment of the magnetic field with respect to the [100] axis as it is scanned. The four classes could for instance be truly resonant only with zero external magnetic field and not when the B field is large.

Fig. \ref{Alignment} shows the change in PL with respect to the external magnetic field amplitude for various misalignments of the magnetic field direction compared to the [100] diamond crystalline axis of the sample CVD-PPM. We can see that the central drop in PL is not critically affected by the misalignment of the magnetic field, and we are therefore confident that the drop in PL we observe when $B\parallel [100]$ is not an effect of misalignment.

The initial alignment of the magnetic field was controlled by applying a field of $\sim 50\ \rm G$ and monitoring the overlap of the four classes with ODMR spectra (maximizing the PL drop is also an effective method). We estimate the initial alignment precision to be $\pm 1^\circ$. The misalignment was then added by rotating the electromagnet along an arbitrary axis. The slight asymmetry between the positive and negative value of the magnetic field could come from the earth magnetic field or from hysteresis in the magnetic core of our electromagnet.

\subsection{Emulation of the electric field with transverse magnetic field}
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_transverse_field}
\caption{(a) Simulated energy level of the three eigenstates $\ket{g},\ket{d}$ and $\ket{e}$ for an NV center under purely transverse magnetic field. The spin Hamiltonian is given in eq. \ref{NV Hamiltonian} and a value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen (b) Blue dashed curve : frequency splitting between the $\ket{d}$ and $\ket{e}$ states as a function of the transverse magnetic field, red plain line : matching factor $\abs{\bra{e}\ket{+}}^2$ between the $\ket{e}$ and $\ket{+}$ states}
\label{theory_transverse_field}
\end{figure}
In the main text we use the property that, under purely transverse magnetic field, the eigenstates of the single spin Hamiltonian are close to the $\{ \ket{0},\ket{+},\ket{-} \}$ basis, described in the main text and in sec. \ref{sec Hamiltonian} of these supplementary. We will discuss here the validity of this claim.

@@

\begin{itemize}
\item The dipole-induced spin decay is lowest for a spectrally isolated class of NV centers when the transverse magnetic field is negligible. In this case the dipole-dipole relaxation is limited by flip-flop processes between spins of the same class. We will call the corresponding stretched lifetime $T_1^{\rm dd}\equiv T_0$.
\item For a single class, still spectrally isolated   from the three other classes, but this time dominated by the electric field or transverse magnetic field (as long as $\gamma_e B_\perp \ll D$), the change in the eigenstates of the single particle Hamiltonian results in an increase of the average flip-flop rate. This leads to a new theoretical stretched lifetime $T_1^{\rm dd}= 4\, T_0$.
\item When a magnetic field is applied along the [100] crystalline axis (for $5\ \rm G < \abs{B} \ll D$), all four classes are resonant and dominated by the longitudinal part of the magnetic field. This results, in increase in the average flip-flop rate compared to the isolated class case, and the predicted stretched exponential lifetime is $T_1^{\rm dd}\approx 42.8\, T_0$.
\item In zero external magnetic field, all four classes are also resonant but we assume that the single particle Hamiltonian of each spin is dominated by a random local electric field. The averaging of flip-flop in this case yields a stretched lifetime $T_1^{\rm dd}\approx 51.4\, T_0$, which is about 20\% higher than the predicted lifetime for $B \parallel$ [100].
\end{itemize}

@@

Assuming a homogeneous distribution of the fluctuators in the bulk of the crystal, the authors of \cite{choi_depolarization_2017} conclude that the fluctuators create an additionnal decay channel for the NV population, through dipole-dipole interaction, characterized by a decay rate $\gamma$ which follows the probability distribution :
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the timescale $T$ is defined as
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f}.
\label{eq 1/T}
\end{equation}
Where $n_f$ is the fluctuator density in the crystal in nm$^{-3}$, $J_0=52\ \rm{MHz}\cdot\rm{nm}^3$ is the characteristic dipole-dipole strength between two spins, $\gamma_f$ is the fluctuator intrinsic decay rate and $\bar \eta$ is a dimension-less number which characterizes the average dipole-dipole interaction between the NV centers and the fluctuators (resonance conditions, relative orientations etc. Further details are given below).

@@

The dipole-dipole Hamiltonian can then be decomposed as :
\begin{align}
-\frac{\mathcal{H}_{\rm dd}}{J_0/r^3}&= 3\left({\bm S}_1 \cdot \hat{u} \right)\left({\bm S}_2 \cdot \hat{u} \right) - {\bm S}_1 \cdot {\bm S}_2  \\
&=\left[3(\hat{u}\cdot\hat{x}_1)(\hat{u}\cdot\hat{x}_2) -\hat{x}_1\cdot\hat{x}_2\right] S_x^1S_x^2 \\
&+\left[3(\hat{u}\cdot\hat{y}_1)(\hat{u}\cdot\hat{y}_2) -\hat{y}_1\cdot\hat{y}_2\right] S_y^1S_y^2 \\
&+\left[3(\hat{u}\cdot\hat{x}_1)(\hat{u}\cdot\hat{y}_2) -\hat{x}_1\cdot\hat{y}_2\right] S_x^1S_y^2 \\
&+\left[3(\hat{u}\cdot\hat{y}_1)(\hat{u}\cdot\hat{x}_2) -\hat{y}_1\cdot\hat{x}_2\right] S_y^1S_x^2 \\
&+\left[3(\hat{u}\cdot\hat{z}_1)(\hat{u}\cdot\hat{z}_2) -\hat{z}_1\cdot\hat{z}_2\right] S_z^1S_z^2, \\
&+\mathcal{H}_{\rm other}
\end{align}
where $J_0=\frac{\mu_0 \gamma_e^2 \hbar^2}{4 \pi} = (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\bm r= r\bm{\hat u}$ is the distance between the two spins and  $\mathcal{H}_{\rm other}$ contains terms of the form $S_{x/y}^iS_z^j$ which couple states far from resonance and will be neglected here.

@@

\begin{equation}
\label{eq. eta bar}
\bar \eta =\frac{1}{4} \sum_{i=1}^4 \int_\theta \int_\phi  \int_{\omega_f} \int_{\omega_{NV}}   \abs{\eta(\bm u,i,\omega_{NV},\omega_f)}\, d\Omega\, \rho(\omega_{NV}) d\omega_{NV}\, \rho^i(\omega_f) d\omega_f 
\end{equation}

Where $d\Omega=\sin \theta d\theta d\phi$, $\rho(\omega_{NV})$ is the distribution of angular frequencies of the probed NV centers, and  $\rho^i(\omega_f)$ is the distribution of angular frequencies  for the fluctuators of class $i$.

@@

With these hypotheses, we find that there are 3 possible scenarios where the NV and fluctuator are resonant :
\begin{itemize}
\item The NV and fluctuator are from the same class, so that the angle $\widehat{z_1 z_2}=0$ : this is always true, regardless of the magnetic field angle and amplitude. We can analytically compute $\bar \eta$ in this case and find : $$ \bar {\eta}_{same}=\frac{1}{4}  \int_\theta \int_\phi \abs{\eta(\bm u,\widehat{z_1 z_2}=0,\omega_{NV}=\omega_f)}\, d\Omega= \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot \frac{2}{3 \sqrt{3}} \approx 5.55\cdot10^{-2}$$
\item The angle $\widehat{z_1 z_2}=\rm{arccos}(\frac{1}{3})\approx 70.5^\circ$. This is the case when the magnetic field lies in the \{100\} crystalline planes family. In this case, numerical simulations yield $\bar \eta_{\rm close} \approx \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot 0.6507 \approx 9.39\cdot10^{-2}$
\item The angle $\widehat{z_1 z_2}=\rm{arccos}(-\frac{1}{3})\approx 109.5^\circ$. This happens when the magnetic field lies in the \{110\} or \{$1\bar{1}0$\} crystalline planes family. In this case, numerically simulations give $\bar \eta_{\rm far} \approx \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot 0.8328 \approx 1.20\cdot10^{-1}$. This last case was not taken into account in \cite{choi_depolarization_2017}.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/Various_ESR}
\caption{ODMR spectra for different orientations of the magnetic field. Red numbers represent the number of degenerate classes for each line of the spectrum}
\label{Various ODMR}
\end{figure}

>>>>

\section{NV$^-$ ground state Hamiltonian under low magnetic field}
\label{sec Hamiltonian}
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_ESR}
\caption{ODMR measurement (a) under zero magnetic field, (b) under a magnetic field of $~ 50 G$ when zooming on a single NV class.}
\label{ESR_single_spin}
\end{figure}
Under zero external magnetic field, there are three possible processes that can cause a splitting of the $\{\ket{+1},\ket{-1}\}$ states : local electric fields, crystal strain and local magnetic fields. Of these three causes, only the electric field can explain the shape of the ODMR line that is observed under zero external magnetic field (Fig. \ref{ESR_single_spin}a) \cite{mittiga2018imaging}.

@@

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/map_etats_propres}
\caption{Numerical simulations showing the closeness of the Hamiltonian second excited state $\ket{e}$ with the states $\ket{+1}$ and $\ket{+}$, as a function of the magnetic field amplitude and angle $\theta$ with respect to the NV axis. A value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen.}
\label{map etats propres}
\end{figure}

In our model for  dipole-dipole coupling under small magnetic fields, we will therefore neglect the contribution of the strain, local magnetic fields and longitudinal electric fields.  We also do not take into account the hyper-fine structure of the NV center because of the large inhomogeneous broadening of the transitions in all our samples (Fig. \ref{ESR_single_spin}b).
We will then consider the following spin Hamiltonian for the NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_s/h=D S_z^2 + \gamma_e \bm{B}_{\rm ext} \cdot \bm{S}+ d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right],
\end{equation}
where $D=2.87\ \rm GHz$ is the zero field splitting and $\gamma_e=2.8\ \rm MHz/G$ the gyromagnetic ratio of the electron.

@@

\section{Samples}
%Note pour moi : j'ai inversé adamas 1 et 2 parraport aux noms des dossiers dans 20220324
Here are the various samples used in this study :
\begin{itemize}
\item \textbf{HPHT-150-1} : A 150 $\mu$m HPHT 1b diamond irradiated and annealed to reach a concentration [NV$^-$] $\approx 3\ \rm ppm$. This sample was bought from Adamas Nanotechnology (MDNV150umHi). It is used in Fig. 2 (b), 2(d) and 3 of the main text, as well as Fig. \ref{ESR_single_spin}, \ref{T1_fits} and \ref{alphas} of the SI.
\item \textbf{HPHT-150-2} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. 4 of the main text and Fig. \ref{largeur_fluct} (c) of the SI.
\item \textbf{HPHT-150-3} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. \ref{Pola} of the SI.
\item \textbf{HPHT-15-1}  : A 15 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV15umHi). This sample is used in Fig. 5 of the main text.
\item \textbf{HPHT-15-2}  : Another sample from the same batch as HPHT-15-1. This sample is used in Fig. \ref{121 VS 22 fig} and \ref{Various ODMR} of the SI.
\item \textbf{HPHT-1-1}  : A 1 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV1umHi). This sample is used to perform a sensitivity measurement similar to the one done in Fig. 5 of the main text.
\item \textbf{CVD-1} : A CVD type IIa bulk diamond. This sample has not been irradiated and contains [NV$^-$] $\approx 50\ \rm ppb$. This sample is used in Fig. 2(a) of the main text.
\item \textbf{CVD-2} : A CVD bulk diamond described in \citep{TALLAIRE2020421}. This one has been irradiated and annealed and contains [NV$^-$] $\approx 4\ \rm ppm$. This sample is used in Fig. \ref{Alignment} of the SI.

@@

\section{Experimental Setup}
\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/shema_exp}
\caption{Experimental Setup.}
\label{setup}
\end{figure}
Fig. \ref{setup} shows the general purpose experimental setup used for all the experiments presented in this article.

@@

We did not use magnetic shielding to protect from the earth magnetic field since the splitting due to  $B_{\rm earth}$ is lower than the splitting due to the local electric field for the samples used here : $ \gamma_e B_{\rm earth} \approx 1.5\ \rm MHz < d_\perp E_\perp \approx 4\ \rm MHz$.

\section{Experimental details}
\subsection{$T_1$ fitting Protocol}
\begin{figure}
\includegraphics[width=0.8\textwidth]{Figures_SI/Fig_T1_combined}
\caption{$T_1$ measurement with purely exponential and purely stretched exponential fits (a) in zero magnetic field (b) in non-zero magnetic field.}
\label{T1_fits}
\end{figure}
\begin{figure}[h]
\includegraphics[width=0.45\textwidth]{Figures_SI/fig_alphas}
\caption{Best stretch factor $\beta$ for a $T_1$ fit of the form $f(\tau)=A \exp(-(\frac{\tau}{T_1})^\beta)$ as a function of an arbitrarely oriented magnetic field amplitude.}
\label{alphas}
\end{figure}

@@

Fig. \ref{T1_fits} shows the result of a lifetime measurement, measured with the subtraction protocol described in the main text, for zero and non-zero magnetic field. The result is fitted with exponential and stretched-exponential (with a stretch factor $\beta=0.5$) functions. We can see that for the shorter lifetime (at zero field), the measurement follows more closely the stretched-exponential profile, while the longer lifetime is closer to the exponential one (except at very short times).

@@

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/overlap}
\caption{Illustration of the spectral overlap for two gaussian spectra.}
\label{overlap}
\end{figure}

@@

As illustrated in Fig. \ref{overlap}, we define the spectral overlap $S(\Delta \nu)$ between two spins of spectral response $S_1(\nu)$ and $S_2(\nu)$, centered respectively on the frequencies $\nu_1$ and $\nu_2$ where $\Delta \nu = \nu_2-\nu_1$ as :
\begin{equation}
S(\Delta \nu)=\int S_1(\nu, \nu_1)S_2(\nu, \nu_2) d\nu.
\end{equation}
In order to approximate the spectral overlap in our experiment, we will consider the analytical solution in the Gaussian and Lorentzian case :

\begin{itemize}
\item For two gaussians of standard deviation $\sigma$, the spectral overlap as a function of the detuning $\Delta \nu = \nu_1-\nu_2$ is itself a gaussian of standard deviation $\sigma'=\sqrt{2} \sigma$. :
\begin{align*}
S(\Delta \nu)&\propto \int \exp(-\frac{(\nu-\nu_1)^2}{2\sigma^2})\exp(-\frac{(\nu-\nu_2)^2}{2\sigma^2}) d\nu \\
&\propto\exp(-\frac{(\Delta \nu)^2}{4\sigma^2}).
\end{align*}

\item For two Lorentzian profile with width $\sigma$, the overlap function is itself a Lorentizan with width $\sigma'=2\sigma$ :
\begin{align*}
S(\Delta \nu)&\propto \int \frac{1}{1+ \frac{(\nu-\nu_1)^2}{\sigma^2}}\cdot \frac{1}{1+ \frac{(\nu-\nu_2)^2}{\sigma^2}} d\nu \\
&\propto\frac{1}{1+ \frac{(\Delta \nu)^2}{4\sigma^2}}.
\end{align*}
\end{itemize}

@@

In particular, we did not observe the apparition of an anti-line inside the main dip, unlike what was observed in the two previously cited work. We expect that the main reason behind the different behaviors is that we seem to observe far greater spin depolarization in zero-field, which we attribute to dipole-dipole coupling. This effect could hide smaller effects such as the one related to the laser polarization.
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_Pola}
\caption{Effect of the polarization of the incident laser. (a) Photoluminescence of sample HPHT-150-3 as a function of arbitrarely oriented magnetic field amplitude for various polarization angle. (b) Demodulated PL in the same conditions.}
\label{Pola}
\end{figure}
\subsection{Alignment of the magnetic field along [100] axis}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{Figures_SI/alignement}
\caption{PL of sample CVD-PPM as a function of the external magnetic field amplitude for different magnetic field misalignments with respect to the [100] axis.}
\label{Alignment}
\end{figure}
We claim that the decrease in PL at low magnetic field in Fig. 3 (c)(ii) of the main text is due to the local electric field and the double flip processes. One other possibility would be a misalignment of the magnetic field with respect to the [100] axis as it is scanned. The four classes could for instance be truly resonant only with zero external magnetic field and not when the B field is large.

Fig. \ref{Alignment} shows the change in PL with respect to the external magnetic field amplitude for various misalignments of the magnetic field direction compared to the [100] diamond crystalline axis of the sample CVD-2. We can see that the central drop in PL is not critically affected by the misalignment of the magnetic field, and we are therefore confident that the drop in PL we observe when $B\parallel [100]$ is not an effect of misalignment.

The initial alignment of the magnetic field was controlled by applying a field of $\sim 50\ \rm G$ and monitoring the overlap of the four classes with ODMR spectra (maximizing the PL drop is also an effective method). We estimate the initial alignment precision to be $\pm 1^\circ$. The misalignment was then introduced by rotating the electromagnet along an arbitrary axis. The slight asymmetry between the positive and negative value of the magnetic field could come from the earth magnetic field or from hysteresis in the magnetic core of our electromagnet.

\subsection{Emulation of the electric field with transverse magnetic field}
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_transverse_field}
\caption{(a) Simulated energy level of the three eigenstates $\ket{g},\ket{d}$ and $\ket{e}$ for an NV center under purely transverse magnetic field. The spin Hamiltonian is given in eq. \ref{NV Hamiltonian} and a value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen (b) Blue dashed curve : frequency splitting between the $\ket{d}$ and $\ket{e}$ states as a function of the transverse magnetic field, red plain line : matching factor $\abs{\bra{e}\ket{+}}^2$ between the $\ket{e}$ and $\ket{+}$ states.}
\label{theory_transverse_field}
\end{figure}
In the main text we use the property that, under purely transverse magnetic field, the eigenstates of the single spin Hamiltonian are close to the $\{ \ket{0},\ket{+},\ket{-} \}$ basis, described in the main text and in sec. \ref{sec Hamiltonian} of these supplementary. We will discuss here the validity of this claim.

@@

\begin{itemize}
\item The dipole-induced spin decay is lowest for a spectrally isolated class of NV centers when the transverse magnetic field is negligible. In this case the dipole-dipole relaxation is limited by flip-flop processes between spins of the same class. We will call the corresponding stretched lifetime $T_1^{\rm dd}\equiv T_0$.
\item For a single class, still spectrally isolated   from the three other classes, but this time dominated by the electric field or transverse magnetic field (as long as $\gamma_e B_\perp \ll D$), the change in the eigenstates of the single particle Hamiltonian results in an increase of the average flip-flop rate. This leads to a new theoretical stretched lifetime $T_1^{\rm dd}= 4\, T_0$.
\item When a magnetic field is applied along the [100] axis (for $5\ \rm G < \abs{B} \ll D/\gamma_e$), all four classes are resonant and dominated by the longitudinal part of the magnetic field. This results, in increase in the average flip-flop rate compared to the isolated class case, and the predicted stretched exponential lifetime is $T_1^{\rm dd}\approx 42.8\, T_0$.
\item In zero external magnetic field, all four classes are also resonant but we assume that the single particle Hamiltonian of each spin is dominated by a random local electric field. The averaging of flip-flop in this case yields a stretched lifetime $T_1^{\rm dd}\approx 51.4\, T_0$, which is about 20\% higher than the predicted lifetime for $B \parallel$ [100].
\end{itemize}

@@

Assuming a homogeneous distribution of the fluctuators in the bulk of the crystal, the authors of \cite{choi_depolarization_2017} show that the fluctuators create an additionnal decay channel for the NV population, through dipole-dipole interaction, characterized by a decay rate $\gamma$ which follows the probability distribution :
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the timescale $T$ is defined as
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f},
\label{eq 1/T}
\end{equation}
where $n_f$ is the fluctuator density in the crystal in nm$^{-3}$, $J_0=52\ \rm{MHz}\cdot\rm{nm}^3$ is the characteristic dipole-dipole strength between two spins, $\gamma_f$ is the fluctuator intrinsic decay rate and $\bar \eta$ is a dimension-less number which characterizes the average dipole-dipole interaction between the NV centers and the fluctuators (resonance conditions, relative orientations etc. Further details are given below).

@@

The dipole-dipole Hamiltonian can then be decomposed as :
\begin{align}
-\frac{\mathcal{H}_{\rm dd}}{J_0/r^3}&= 3\left({\bm S}_1 \cdot \hat{u} \right)\left({\bm S}_2 \cdot \hat{u} \right) - {\bm S}_1 \cdot {\bm S}_2  \\
&=\left[3(\hat{u}\cdot\hat{x}_1)(\hat{u}\cdot\hat{x}_2) -\hat{x}_1\cdot\hat{x}_2\right] S_x^1S_x^2 \\
&+\left[3(\hat{u}\cdot\hat{y}_1)(\hat{u}\cdot\hat{y}_2) -\hat{y}_1\cdot\hat{y}_2\right] S_y^1S_y^2 \\
&+\left[3(\hat{u}\cdot\hat{x}_1)(\hat{u}\cdot\hat{y}_2) -\hat{x}_1\cdot\hat{y}_2\right] S_x^1S_y^2 \\
&+\left[3(\hat{u}\cdot\hat{y}_1)(\hat{u}\cdot\hat{x}_2) -\hat{y}_1\cdot\hat{x}_2\right] S_y^1S_x^2 \\
&+\left[3(\hat{u}\cdot\hat{z}_1)(\hat{u}\cdot\hat{z}_2) -\hat{z}_1\cdot\hat{z}_2\right] S_z^1S_z^2, \\
&+\mathcal{H}_{\rm other},
\end{align}
where $J_0=\frac{\mu_0 \gamma_e^2 \hbar^2}{4 \pi} = (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\bm r= r\bm{\hat u}$ is the distance between the two spins and  $\mathcal{H}_{\rm other}$ contains terms of the form $S_x^iS_z^j$ and $S_y^iS_z^j$ which couple states far from resonance and will be neglected here.

@@

\begin{equation}
\label{eq. eta bar}
\bar \eta =\frac{1}{4} \sum_{i=1}^4 \int_\theta \int_\phi  \int_{\omega_f} \int_{\omega_{NV}}   \abs{\eta(\bm u,i,\omega_{NV},\omega_f)}\, d\Omega\, \rho(\omega_{NV}) d\omega_{NV}\, \rho^i(\omega_f) d\omega_f ,
\end{equation}

where $d\Omega=\sin \theta d\theta d\phi$, $\rho(\omega_{NV})$ is the distribution of angular frequencies of the probed NV centers, and  $\rho^i(\omega_f)$ is the distribution of angular frequencies  for the fluctuators of class $i$.

@@

With these hypotheses, we find that there are 3 possible scenarios where the NV and fluctuator are resonant :
\begin{itemize}
\item The NV and fluctuator are from the same class, so that the angle $\widehat{z_1 z_2}=0$ : this is always true, regardless of the magnetic field angle and amplitude. We can analytically compute $\bar \eta$ in this case and find : $$ \bar {\eta}_{same}=\frac{1}{4}  \int_\theta \int_\phi \abs{\eta(\bm u,\widehat{z_1 z_2}=0,\omega_{NV}=\omega_f)}\, d\Omega= \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot \frac{2}{3 \sqrt{3}} \approx 5.55\cdot10^{-2}.$$
\item The angle $\widehat{z_1 z_2}=\rm{arccos}(\frac{1}{3})\approx 70.5^\circ$. This is the case when the magnetic field lies in the \{100\} crystalline planes family. In this case, numerical simulations yield $\bar \eta_{\rm close} \approx \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot 0.6507 \approx 9.39\cdot10^{-2}$
\item The angle $\widehat{z_1 z_2}=\rm{arccos}(-\frac{1}{3})\approx 109.5^\circ$. This happens when the magnetic field lies in the \{110\} or \{$1\bar{1}0$\} crystalline planes family. In this case, numerically simulations give $\bar \eta_{\rm far} \approx \frac{1}{4} \cdot \sqrt{\frac{1}{3}} \cdot 0.8328 \approx 1.20\cdot10^{-1}$. This last case was not taken into account in \cite{choi_depolarization_2017}.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/Various_ESR}
\caption{ODMR spectra for different orientations of the magnetic field. Red numbers represent the number of degenerate classes for each line of the spectrum.}
\label{Various ODMR}
\end{figure}
